UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts

### 2019-07-08 ms/main.tex
suggested edit to FEV1

A few individuals cluster with populations different from their label: some Mexican individuals cluster with Spanish and Italian populations; some Puerto Rican individuals cluster with African American and Caribbean populations; and one Gambian individual clusters with the Mende of Sierra Leone. Two populations form multiple clusters: Gujarati Indians in Houston, Texas and Punjabi people in Lahore, Pakistan. This clustering effect is robust to the number of components considered (figure~\ref{fig:supp_megamontage_pc10_50}). Differentiation in the Gujarati population from substructure has been previously noted\cite{reich2009india}. Recontacting of 23andMe participants associated with the clusters show that one of the cluster is highly enriched in individuals from a group in Western India with shared ancestry and patronym. [D. Poznik, 23andme, personal communication, and \cite{23andme}].

@@

\subsubsection*{Population structure in the UKBB reflects local and global genetic variation} 
The UKBB contains data on 488,377 individuals containing phenotypic measures and self-identified ethnic backgrounds. Figure~\ref{fig:fig_ukbb} compares UMAP to PCA apply to UKBB. As expected, PCA captures major axes of variation emphasizing continental ancestry, whereas UMAP reveals finer sturcture. UMAP on the top 10 principal components reveals continuous and discrete population structure (figure~\ref{fig:umap_ukbb_pc10}): the patchwork of local topologies identifies multiple sub-populations, as well as continuous structure within populations and admixture gradients between populations. The result is a comprehensive portrait of genetic variation capturing population relationships not visible using other methods, succinctly illustrating the complex structure of a large and multi-ethnic dataset.

@@

\adpcomment{Discussing FEV1 here. Note: citations tend to present relative differences b/w groups as a percentage so I used UKBB data.}
Figure~\ref{fig:supp_comparison_fev_chi_eur} further suggests a difference in FEV1 between those who self-identified as Chinese and a nearby cluster enriched in individuals born in Japan; to our knowledge there have not been studies into differences in FEV1 between these populations. Baseline FEV1 levels are determined via equations that include ethnicity or race\cite{quanjer2012multi}, but studies in several populations have shown that there is considerable variation based on ancestry, even within self-defined ethnicity\cite{ortega2015effect}. We selected the individuals whose UMAP coordinates were near the Chinese cluster and kept those born in Japan, Malaysia, and the Philippines as well as the self-identified Chinese population. These four groups are mutually exclusive and are shown in figure~\ref{fig:supp_fev_test_pops}. Using this approach facilitates identifying large phenotypic differences among genetically related groups, which is useful when population labels are unavailable or correlate poorly with genetic ancestry. After adjusting for age, age$^2$, height, and sex, an unpaired two-sample t-test shows those born in Japan have a higher mean FEV1 than Chinese individuals by 0.224 (p-value $2.787\times 10^{-15}$). By sex, there is a difference of 0.213 (p-value $5.4\times 10^{-13}$) among females and 0.317 (p-value $5.1\times 10^{-4}$) among males, though there are considerably fewer males in the sample (distributions presented in figure~\ref{fig:supp_fev_ridgeplots}). For comparison, the adjusted difference between self-identified African and British individuals in the UKBB is 0.762 (p-value $2.2\times 10^{-16}$).

>>>>

A few individuals cluster with populations different from their label: some Mexican individuals cluster with Spanish and Italian populations; some Puerto Rican individuals cluster with African American and Caribbean populations; and one Gambian individual clusters with the Mende of Sierra Leone. Two populations form multiple clusters: Gujarati Indians in Houston, Texas and Punjabi people in Lahore, Pakistan. This clustering effect is robust to the number of components considered (figure~\ref{fig:supp_megamontage_pc10_50}). Differentiation in the 1KGP Gujarati population has been previously identified through a PCA restricted to the Gujarati \cite{reich2009india}. Following a preprint version of the present article, 23andme released a statement \cite{23andme} arguing that one of the two clusters could be traced, via 23andMe participant recontact to individuals from a group in Western India with shared ancestry and patronym. [D. Poznik, 23andme, personal communication, and \cite{23andme}].

@@

\subsubsection*{Population structure in the UKBB reflects local and global genetic variation} 
The UKBB contains data on 488,377 individuals including genotypes, phenotypic measures and self-identified ethnic backgrounds. Figure~\ref{fig:fig_ukbb} compares UMAP to PCA applied to UKBB. As expected, PCA captures major axes of variation emphasizing continental ancestry, whereas UMAP reveals finer sturcture. UMAP on the top 10 principal components reveals continuous and discrete population structure (figure~\ref{fig:umap_ukbb_pc10}): the patchwork of local topologies identifies multiple sub-populations, as well as continuous structure within populations and admixture gradients between populations. The result is a comprehensive portrait of genetic variation capturing population relationships not visible using other methods, succinctly illustrating the complex structure of a large and multi-ethnic dataset.

@@

\adpcomment{Discussing FEV1 here. Note: citations tend to present relative differences b/w groups as a percentage so I used UKBB data.}
Figure~\ref{fig:supp_comparison_fev_chi_eur} further suggests a difference in FEV1 between those who self-identified as Chinese and a nearby cluster enriched in individuals born in Japan; to our knowledge there have not been studies into differences in FEV1 between these populations. Baseline FEV1 levels are determined via equations that include ethnicity or race\cite{quanjer2012multi}, but studies in several populations have shown that there is considerable variation based on ancestry, even within self-defined ethnicity\cite{ortega2015effect}. To focus on individuals of Asian ancestry (rather than, e.g., individuals born in Japan but who have European ancestry), we first selected the individuals whose UMAP coordinates were near the Chinese cluster. We then focused on individuals born in Japan, Malaysia, and the Philippines as well as the self-identified Chinese population. These four groups are mutually exclusive and are shown in figure~\ref{fig:supp_fev_test_pops}. Using this approach facilitates identifying large phenotypic differences among genetically related groups, which is useful when population labels are unavailable or correlate poorly with genetic ancestry. After adjusting for age, age$^2$, height, and sex, an unpaired two-sample t-test shows those born in Japan have a higher mean FEV1 than Chinese individuals by 0.224 (p-value $2.787\times 10^{-15}$). By sex, there is a difference of 0.213 (p-value $5.4\times 10^{-13}$) among females and 0.317 (p-value $5.1\times 10^{-4}$) among males, though there are considerably fewer males in the sample (distributions presented in figure~\ref{fig:supp_fev_ridgeplots}). For comparison, the adjusted difference between self-identified African and British individuals in the UKBB is 0.762 (p-value $2.2\times 10^{-16}$).
